---
title: "Interpretable Machine Learning projekt ACB"
author: "Anton Bendsen"
date: "2026-04-30"
output: html_document

NOTE: Depends om the amount of cores.

```{r}
library(future)
future::plan(multisession, workers = 8)
```



```{r}
library(mlr3verse)
```


```{r}
#install.packages("CASdatasets", repos = "http://cas.uqam.ca/pub/", type="source")
```



```{r}
library(CASdatasets)

data(freMPL1) 

data(freMPL2) 
data(freMPL3) 

data(freMPL4)

freMPL3 <- subset( freMPL3 , select = -DeducType )  

freMPL4 <- subset( freMPL4 , select = -DeducType )  

freMPL <- rbind(freMPL1,freMPL2,freMPL3,freMPL4)


```

alternative: (has X col.)
```{r}
#freMPLabsalon <- read.csv("~/Downloads/freMPL.csv")
```


We remove the col "RecordEnd": and altså claim indicator.

```{r}
freMPL$HasRecordEnd <- ifelse(is.na(freMPL$RecordEnd), 0, 1)

freMPLnoUNA <- subset(freMPL, select = -RecordEnd)
freMPLnoUNA <- subset(freMPLnoUNA, select = -ClaimInd)
freMPLnoUNA$AmountExposure=freMPLnoUNA$ClaimAmount/freMPLnoUNA$Exposure

```

We see that we have no NA's now:
```{r}
nrow(freMPLnoUNA)
nrow(na.omit(freMPLnoUNA))
```


We see that much of the data are dublicates. This seem very unrealistic. We therefore remove all
```{r}
unique(freMPLnoUNA)
```

```{r}
freMPLnoUNA<-unique(freMPLnoUNA)
```


To handle SocialCategories:

```{r}
library(dplyr)

freMPLnoUNA$SocioGroup <- dplyr::case_when(
  grepl("^CSP1", freMPLnoUNA$SocioCateg) ~ "Farmer",
  grepl("^CSP2", freMPLnoUNA$SocioCateg) ~ "SelfEmp",
  grepl("^CSP3", freMPLnoUNA$SocioCateg) ~ "Executive",
  grepl("^CSP4", freMPLnoUNA$SocioCateg) ~ "Intermediate",
  grepl("^CSP5", freMPLnoUNA$SocioCateg) ~ "Employee",
  grepl("^CSP6", freMPLnoUNA$SocioCateg) ~ "Worker",
  grepl("^CSP7", freMPLnoUNA$SocioCateg) ~ "Retired",
  grepl("^CSP8", freMPLnoUNA$SocioCateg) ~ "Inactive",
  TRUE ~ "Unknown"
)

table(freMPLnoUNA$SocioGroup)

freMPLnoUNA$SocioGroup<-as.factor(freMPLnoUNA$SocioGroup)
```
from https://fr.wikipedia.org/wiki/Professions_et_cat%C3%A9gories_socioprofessionnelles_en_France


```{r}
freMPLnoUNA <- subset(freMPLnoUNA, select = -SocioCateg)

summary(freMPLnoUNA)
```




```{r}
freMPLnoUNA$RecordBeg <- as.numeric(freMPLnoUNA$RecordBeg)-12418 #12418=1/1-2004, thus RecordBeg will be days after 1/1-2004.
```

If the ClaimAmount is negative we set it to zero:

```{r}
freMPLnoUNA$AmountExposure[freMPLnoUNA$AmountExposure < 0] <- 0

freMPLnoUNA<-subset(freMPLnoUNA,select = -ClaimAmount)


```




# make task
```{r}
claim_task=as_task_regr(freMPLnoUNA,id="claim_task",target="AmountExposure",weight = freMPLnoUNA$Exposure)


claim_task$select(
  setdiff(
    claim_task$feature_names,
    "Exposure"
  )
)

```

We dont remove then.

Fit first model:
```{r}
graph1 <- 
  po("encode") %>>%
  po("scale") %>>%
  lrn("regr.lm")

graph1_learner <- as_learner(graph1)
graph1$plot(horizontal = TRUE)

```

We split to have a validation set:

```{r}
set.seed(123)
splits = partition(claim_task,ratio=0.8)

task_train = claim_task$clone()$filter(splits$train)
task_test  = claim_task$clone()$filter(splits$test)
```

```{r}
#graph1_learner$train(claim_task,row_ids=splits$train)

graph1_learner$train(task_train)

graph1_learner$model
```


```{r}
#pred = graph1_learner$predict(claim_task, row_ids = splits$test)
pred = graph1_learner$predict(task_test)


pred

pred$score(msr("regr.mse"))
pred$score(msr("regr.mae"))

```

The above is for a single fold. We would like to do CV to get stronger results:

```{r}
resampling = rsmp("cv", folds = 10)

rr_graph1 = resample(
  task_train,
  graph1_learner,
  resampling
)

rr_graph1$aggregate(msr("regr.mse"))
rr_graph1$aggregate(msr("regr.mae"))
```

If we would see info about the model:
```{r}
graph1_learner$model$regr.lm$model

rr_graph1$model$regr.lm$model
```





Compare to featureless:

```{r} 
##claim_task$set_col_roles("Exposure", roles = "weights_learner") (virker ikke)

graphfeatureless <- 
  po("encode") %>>%
#  po("scale") %>>%
  lrn("regr.featureless")

graphfeatureless_learner <- as_learner(graphfeatureless)

graphfeatureless_learner$train(claim_task,row_ids=splits$train)

#predfeatureless = graphfeatureless_learner$predict(claim_task, row_ids = splits$test)
predfeatureless = graphfeatureless_learner$predict(task_test)
predfeatureless
predfeatureless$score(msr("regr.mse"))
predfeatureless$score(msr("regr.mae"))



graphfeatureless_learner$predict(task_test)
```


With CV:
```{r}
set.seed(123)

rr_featureless = resample(
  task_train,
  graphfeatureless_learner,
  resampling
)

rr_featureless$aggregate(msr("regr.mse"))
rr_featureless$aggregate(msr("regr.mae"))
```


unique(freMPL2)

To validate (only as the very last):
```{r}
pred = graphfeatureless_learner$predict(task_test)
pred$score(msr("regr.mse"))
```



#Lasso and Ridge regression:


```{r}
LassoGraph <- 
  po("encode") %>>%
  po("scale") %>>%
  lrn("regr.glmnet",id="lasso", alpha=1,lambda=to_tune(p_dbl(log(1e-5), log(1e1), trafo = exp)))

RidgeGraph <- 
  po("encode") %>>%
  po("scale") %>>%
  lrn("regr.glmnet",id="ridge", alpha=0,lambda=to_tune(p_dbl(log(1e-5), log(1e1), trafo = exp)))

ElasticNetGraph <- 
  po("encode") %>>%
  po("scale") %>>%
  lrn("regr.glmnet",id="elasticnet", alpha = to_tune(0, 1),lambda=to_tune(p_dbl(log(1e-5), log(1e1), trafo = exp)))


LassoLearner<-as_learner(LassoGraph)
RidgeLearner<-as_learner(RidgeGraph)
ElasticNetLearner<-as_learner(ElasticNetGraph)


```

Use autotuner:

```{r}
at_Lasso = auto_tuner(
  tuner = tnr("random_search", batch_size = 20),
  learner = LassoLearner,
  resampling = rsmp("cv", folds = 5),
  measure = msr("regr.mse"),
  terminator = trm("evals", n_evals = 50)
)

at_Ridge = auto_tuner(
  tuner = tnr("random_search", batch_size = 20),
  learner = RidgeLearner,
  resampling = rsmp("cv", folds = 5),
  measure = msr("regr.mse"),
  terminator = trm("evals", n_evals = 50)
)

at_ElasticNet = auto_tuner(
  tuner = tnr("random_search", batch_size = 20),
  learner = ElasticNetLearner,
  resampling = rsmp("cv", folds = 5),
  measure = msr("regr.mse"),
  terminator = trm("evals", n_evals = 50)
)
```

```{r}
learners=list(graph1,graphfeatureless,at_Lasso,at_Ridge,at_ElasticNet)

design=benchmark_grid(task=claim_task,learners=learners,resamplings = rsmp("cv",folds=5))
```


```{r}
results_benchmark<-benchmark(design)
```


```{r}
results_benchmark$aggregate(list(msr("regr.mse"),msr("regr.mae")))
```


# WE now try with splines:

```{r}
SplinesLearner<-as_learner(lrn(
    "regr.gam",
    formula=AmountExposure ~ s(BonusMalus, bs="cr") +s(LicAge,bs="cr")+s(RecordBeg,bs="cr")+VehAge+Gender+MariStat+VehUsage+s(DrivAge,bs="cr")+HasKmLimit+VehBody+VehPrice+VehEngine+VehEnergy+VehMaxSpeed+VehClass+s(RiskVar,bs="cr")+Garage+HasRecordEnd+SocioGroup ,
    gamma = to_tune(1,1e3,logscale=TRUE),
    select = to_tune(c(TRUE, FALSE)),
    drop.unused.levels=FALSE # Default is TRUE, but this may cause errors if splits are unlucky
  ))


at_Splines = auto_tuner(
  tnr("random_search", batch_size=20),
  SplinesLearner,
  measure = msr("regr.mse"),
  resampling = rsmp("cv", folds=5),
  terminator = trm("evals", n_evals=50)
)


learner_spline=list(at_Splines)

design_splines=benchmark_grid(task=claim_task,learners=learner_spline,resamplings = rsmp("cv",folds=5))

results_splines<-benchmark(design_splines)

results_splines$aggregate(list(msr("regr.mse"),msr("regr.mae")))

```


To get beta-coef we train on one set:
```{r}
at_Splines$train(claim_task)

best<-at_Splines$tuning_result$learner_param_vals[[1]]

final_gam <- as_learner(po("encode") %>>%
  po("scale") %>>%lrn(
  "regr.gam",
  formula = AmountExposure ~ s(BonusMalus, bs="cr") +
    s(LicAge,bs="cr") +
    s(RecordBeg,bs="cr") +
    VehAge + Gender + MariStat + VehUsage +
    s(DrivAge,bs="cr") + HasKmLimit + VehBody +
    VehPrice + VehEngine + VehEnergy + VehMaxSpeed +
    VehClass + s(RiskVar,bs="cr") + Garage +
    HasRecordEnd + SocioGroup,
  gamma = best$gamma,
  select = best$select,
  drop.unused.levels = FALSE
))

final_gam$train(claim_task)

summary(final_gam$model)
```


#We make a tree:
```{r}
TreeLrn <- lrn("regr.rpart",id="Tree",cp=0,predict_type="response") #CP=0 means no pruning

```

```{r}
learner_tree=list(as_learner(TreeLrn))

design_tree=benchmark_grid(task=claim_task,learners=learner_tree,resamplings = rsmp("cv",folds=5))

results_tree<-benchmark(design_tree)

results_tree$aggregate(list(msr("regr.mse"),msr("regr.mae")))
```
We plot the tree

```{r}
TreeLrn$train(claim_task)

tree_model <- TreeLrn$graph_model$model$rpart$model

TreeLrn$model

rpart::printcp(TreeLrn)



```

```{r}
sum(TreeLrn$model$frame$var != "<leaf>")
```
Vi får altså over 2200 blade.

det er lort.

We add some cp to make it smaller:


```{r}
TreeLrn2 <- lrn("regr.rpart",id="Tree",cp=0.01,predict_type="response") #CP=0 means no pruning
TreeLrn2$train(claim_task)
TreeLrn2$model

```
We see men make losses of 227€ where women make losses of 160.000€...


```{r}
TreeLrn3 <- lrn("regr.rpart",id="Tree",cp=0.001,predict_type="response") #CP=0 means no pruning
TreeLrn3$train(claim_task)
TreeLrn3$model
```

```{r}
TreeTune <- lrn("regr.rpart",id="Tree",predict_type="response",cp = to_tune(p_dbl(1e-4, 0.05, logscale = TRUE))) #CP=0 means no pruning

TreeLrnTune=as_learner(TreeTune)

at_TreeTune = auto_tuner(
  tuner = tnr("grid_search", resolution = 30),
  learner = TreeLrnTune,
  resampling = rsmp("cv", folds = 5),
  measure = msr("regr.mse"),
  terminator = trm("none")
)

learner_Tree=list(at_TreeTune)
design_Tree=benchmark_grid(task=claim_task,learners=learner_Tree,resamplings = rsmp("cv",folds=5))
results_benchmark_Tree<-benchmark(design_Tree)

```

```{r}
results_benchmark_Tree$aggregate(list(msr("regr.mse"),msr("regr.mae")))

```





# Overview of the data set:


```{r}
barplot(proportions(table(5*round(freMPLnoUNA$LicAge/60), freMPLnoUNA$RiskVar), 2), beside = FALSE, 
        col=rainbow(17), xlab="RiskVar", main="LicAge (year)", xlim=c(1, 30))
legend("right", fill=rainbow(17), legend=sort(unique(5*round(freMPLnoUNA$LicAge/60))), cex=0.8)
```

```{r}
lic_age_group <- 5 * round(freMPLnoUNA$LicAge / 60)
n <- length(unique(lic_age_group))

barplot(
  proportions(table(lic_age_group, freMPLnoUNA$RiskVar), 2),
  beside = FALSE,
  col = rainbow(n),
  xlab = "RiskVar",
  main = "LicAge (year)"
)

legend(
  "right",
  fill = rainbow(n),
  legend = sort(unique(lic_age_group)),
  cex = 0.8
)
```



Only looking at the data where we have claims:

```{r}
claim_data<-subset(freMPLnoUNA, ClaimAmount>0)

```









